---
title: "ESM 244 - Lab 1"
author: "Karina Johnston"
date: "1/5/2022"
output: html_document
---

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# attach packages
library(tidyverse)
library(tmap)
library(here)
library(sf)

# install.packages("tmap") - new package
```

## First lab of the quarter!

### Read in the data

```{r}

sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"),
                     show_col_types = FALSE)

# names(sf_trees) - gives column names
# View(sf_trees) - pulls up dataframe
```

### Part 1: wrangling and ggplot review 

**Example 1.** Find counts of observation by `legal_status` and wrangle

```{r}
# method 1: group_by() and summarize()

sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())


# method 2: different way, plus a few new functions

top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% # remember can type ?drop_na in the console to see
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% # to help reorder columms (relocate to front)!!
  slice_max(tree_count, n = 5) %>% 
  arrange(-tree_count) # descending tree count or desc(tree_count)

```

## Make a graph of the top 5 from above using ggplot

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, -tree_count), 
                                y = tree_count)) +
  geom_col(fill = 'darkgreen') +
  theme_classic() +
  labs(x = "Legal Status", y = "Tree Count") +
  coord_flip()

```

## More wrangling

**Example. 2:** only going to keep observations where legal status is "permitted site" and caretaker is "MTA" and store as `permitted_data_df`

```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)

permitted_data_df <- sf_trees %>% 
  filter(legal_status == "Permitted Site", caretaker == "MTA") # could also use an "&"

```

**Example 3.** only keep blackwood acacia trees and only columns legal_status, latitude and longitude

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% # to check for the string
  select(legal_status, date, lat = latitude, lon = longitude)

```

#### Make a graph of locations

```{r}
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = 'darkgreen') +
  theme_classic()

```



